### 2016-10-19 Manuscript/SW_early_life_telomeres.Rmd
Edits

2. The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system to the causes and consequences of individual variation in telomere dynamics. Here we combine this long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years. We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.

@@

5. Our comprehensive, long-term data suggest that although telomeres shorten with age, in natural populations telomere dynamics vary enormously over both space and time, and that this variation cannot be attributed entirely to measurment error. Ascertaining what explains this variation will be a difficult but worthwhile task, and will require combining long-term ecological and quantitative genetic data.

@@

###Introduction
Exposure to favourable environmental conditions during development and growth can confer fitness advantages later in life [so called 'silver spoon effects'; @Grafen1988; @Monaghan2008]. Adult fitness can be affected by a range of early-life experiences, including the quality of the habitat and available resources [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. We expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006]. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and the resulting adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. We therefore need to understand, or at least be able to measure, how the environment affects individuals differentially within a population in order to elucidate the later-life consequences of early-life experience.

Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem. Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

@@

#####Telomere dynamics and age
Using the cross-sectional data RTL decreased with age across the entire lifespan of the Seychelles warbler (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`). We found that modelling a non-linear effect age explained almost double the variation in telomere length (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`. Also worth noting is that while the overall trend was for decreasing delomre length, there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1A). RTL also decreased with age within the first year of life (i.e. excluding adult samples; t = `r round2(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,5])`). In the longitudinal analysis RTL decreased with age within individuals (i.e. with $\Delta$age; t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P = `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig. 1B).

@@

To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we used the entire Seychelles warbler telomere dataset to compare variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 4C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`). We found no evidence that the likelihood of telomere lengthening was related to the follow up time between sampling events (GLM: t = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,3])`; P = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,4])`), and considering those samples that did show increases in $\Delta$RTL, the degree of increase was not related to follow up time (LM: t = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,3])`; P = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,4])`).

>>>>

2. The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system for disentangling the causes and consequences of individual variation in telomere dynamics. Here we study spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler. We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(dd$CatchYear)-min(dd$CatchYear)` years. We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.

@@

5. Our comprehensive, long-term data show that in natural populations telomere dynamics vary enormously over both space and time, and that this variation cannot be attributed entirely to measurment error. Ascertaining what explains this variation will be a difficult but worthwhile task, and will require combining long-term ecological and quantitative genetic data.

@@

###Introduction
Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem. Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

@@

#####Telomere dynamics and age
Using the cross-sectional data RTL decreased with age across the entire lifespan of the Seychelles warbler (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`). We found that modelling a non-linear effect age explained almost double the variation in telomere length (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`. Also worth noting is that while the overall trend was for decreasing telomere length, there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1A).

@@

To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 4C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`). We found no evidence that the likelihood of telomere lengthening was related to the follow up time between sampling events (GLM: t = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,3])`; P = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,4])`), and considering those samples that did show increases in $\Delta$RTL, the degree of increase was not related to follow up time (LM: t = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,3])`; P = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,4])`).
